---
title: "Class 19: Pertussis Mini Project"
author: "Eric Jordahl"
format: pdf
date: 2022-11-30
toc: TRUE
---
# 1. Investigating Pertussis Cases By Year

## Q1
**With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time**
```{r}
#use datapasta to paste copied CDC data table and paste it as dataframe.
cdc <- data.frame(
                          Year = c(1922L,
                                   1923L,1924L,1925L,1926L,1927L,1928L,
                                   1929L,1930L,1931L,1932L,1933L,1934L,1935L,
                                   1936L,1937L,1938L,1939L,1940L,1941L,
                                   1942L,1943L,1944L,1945L,1946L,1947L,1948L,
                                   1949L,1950L,1951L,1952L,1953L,1954L,
                                   1955L,1956L,1957L,1958L,1959L,1960L,
                                   1961L,1962L,1963L,1964L,1965L,1966L,1967L,
                                   1968L,1969L,1970L,1971L,1972L,1973L,
                                   1974L,1975L,1976L,1977L,1978L,1979L,1980L,
                                   1981L,1982L,1983L,1984L,1985L,1986L,
                                   1987L,1988L,1989L,1990L,1991L,1992L,1993L,
                                   1994L,1995L,1996L,1997L,1998L,1999L,
                                   2000L,2001L,2002L,2003L,2004L,2005L,
                                   2006L,2007L,2008L,2009L,2010L,2011L,2012L,
                                   2013L,2014L,2015L,2016L,2017L,2018L,
                                   2019L),
Cases = c(107473,
                                   164191,165418,152003,202210,181411,
                                   161799,197371,166914,172559,215343,179135,
                                   265269,180518,147237,214652,227319,103188,
                                   183866,222202,191383,191890,109873,
                                   133792,109860,156517,74715,69479,120718,
                                   68687,45030,37129,60886,62786,31732,28295,
                                   32148,40005,14809,11468,17749,17135,
                                   13005,6799,7717,9718,4810,3285,4249,
                                   3036,3287,1759,2402,1738,1010,2177,2063,
                                   1623,1730,1248,1895,2463,2276,3589,
                                   4195,2823,3450,4157,4570,2719,4083,6586,
                                   4617,5137,7796,6564,7405,7298,7867,
                                   7580,9771,11647,25827,25616,15632,10454,
                                   13278,16858,27550,18719,48277,28639,
                                   32971,20762,17972,18975,15609,18617)
)
head(cdc)
```

```{r}
library(ggplot2)
ggplot(data = cdc) + aes(x=Year, y=Cases) +
  geom_point() +
  geom_line() +
  labs(title = "Pertussis cases per year in the USA")

baseplot <- ggplot(data = cdc) + aes(x=Year, y=Cases) +
  geom_point() +
  geom_line() +
  labs(title = "Pertussis cases per year in the USA")
```

# 2. A Tale of Two Vaccines

##Q2
**Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?**

```{r}
library(ggplot2)
ggplot(data = cdc) + aes(x=Year, y=Cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, color="RED", linetype = "dotted") +
  geom_vline(xintercept = 1996, color="BLUE", linetype = "dotted") +
  labs(title = "Pertussis cases per year in the USA")

baseplot <- ggplot(data = cdc) + aes(x=Year, y=Cases) +
  geom_point() +
  geom_line() +
  labs(title = "Pertussis cases per year in the USA")
```

## Q3
**Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?**

After the introduction of the aP vaccine, cases seemed to increase slowly, but not to as large of numbers as before the wP vaccine. A possible explanation for this trend is the resistance to administering vaccines that has occurred more recently, or the new aP vaccine is less efficient at preventing pertussis than the wP vaccine.

# 3. Exploring CMI-PB data

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
head(subject, 3)
```
## Q4
**How many aP and wP infancy vaccinated subjects are in the dataset?**

```{r}
table(subject$infancy_vac)
```

There are 47 subjects in the aP dataset and 49 in the wP dataset

## Q5
**How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

There are 66 female subjescts and 30 male subjects

## Q6
**What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?**

```{r}
table(subject$race, subject$biological_sex)
```

## Working with Dates

```{r}
library(lubridate)
today()
```

```{r}
library(lubridate)
today() - ymd("2000-01-01")
```

```{r}
library(lubridate)
time_length( today() - ymd("2000-01-01"),  "years")
```

## Q7
**Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?**

```{r}
subject$age <- today() - ymd(subject$year_of_birth)
```

```{r}
library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )
```

```{r}
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```

## Q8
**Determine the age of all individuals at time of boost?**

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

## Q9
**With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?**

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) 
```

## Joining multiple tables

```{r}
# Complete the API URLs...
specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector = TRUE)
head(specimen)
```

## Q9
**Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details**

```{r}
meta <- inner_join(specimen, subject)
dim(meta)
```

## Q10
**Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.**

```{r}
titer <- read_json("http://cmi-pb.org/api/ab_titer", simplifyVector = TRUE)
head(titer)
```

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```

## Q11
**How many specimens (i.e. entries in abdata) do we have for each isotype?**

```{r}
table(abdata$isotype)
```

## Q12
**What do you notice about the number of visit 8 specimens compared to other visits?**

```{r}
table(abdata$visit)
```

The eighth visit had way less specimens in comparison to the other 7 visits.

# 4. Examine IgG1 Ab titer levels

## Q13
**Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens**

```{r}
ig1 <- abdata %>% filter(isotype == "IgG1", visit!=8)
head(ig1)
```

```{r}
ggplot(ig1) +
  aes(MFI, antigen, color= antigen) +
  geom_boxplot() +
  facet_wrap(vars(visit), nrow=2) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```

## Q14
**What antigens show differences in the level of IgG1 antibody titers recognizing them over time? Why these and not others?**

FIM2/3 appears to have the most different titer level for these antigens. This could be because it could be the only immunodominant antigen, or it could be more accessible than other antigens.

```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  theme_bw()
```

```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```

## Q15
**Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“Measles”, that is not in our vaccines) and a clear antigen of interest (“FIM2/3”, extra-cellular fimbriae proteins from B. pertussis that participate in substrate attachment).**
```{r}
filter(ig1, antigen=="Measles") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = F) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(ig1, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = F) +
  facet_wrap(vars(visit)) +
  theme_bw()
```
