---
title: "Class 05: Data Visualization with ggplot2"
author: "Eric Jordahl"
format: pdf
toc: true
---


# Our first plot

R has base graphics

```{r}
plot(cars)
```

How would I plot this with `ggplot2`?

Every ggplot nees at least 3 layers:

- **Data** (i.e. the data.frame we have),

- **Aes** (the aesthettic mapping of our data to what we want to plot)

- **Geoms** (How we want to plot this stuff!)

```{r}
library(ggplot2)
ggplot(data=cars) + aes(x=speed, y=dist) + geom_point() + labs(title= "Car speed vs Stopping Distance", x = "Speed (MPH)", y = "Stopping Distance (Feet)", subtitle =  "Eric Jordahl", caption = "Dataset: 'cars'") + geom_smooth(method=lm, se=FALSE) + theme_bw()
```

# A More Interesting Plot

First read the data from online
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

**Q1. How many genes are in this data set?**

There are `r nrow(genes)` genes in this data set.

**Q2. How many columns are in this data set?**

There are `r ncol(genes)` columns in this data set with names `r colnames(genes)`

**Q3. How many upregulated genes are there?**

```{r}
up_genes <- table(genes$State=="up")
up_genes[2]
```
The number of upregulated genes is `r up_genes[2]`

**Q4. What fraction of total genes is up-regulated in this data set?**

The fraction of total genes upregulated is `r up_genes[2]/nrow(genes)*100`


```{r}
library(ggplot2)
p <- ggplot(data=genes) + 
  aes(x=Condition1, y=Condition2, col=State) + 
  geom_point(alpha=0.3) + 
  labs(title="Gene Expression Changes Upon Drug Treatment", x="Control (No Drug)", y="Drug Treatment", subtitle="Eric Jordahl", caption="Dataset: Genes") 
p + scale_colour_manual( values=c("skyblue","gray","orchid") )

```

