---
title: "Class 06: R Functions"
author: "Eric Jordahl"
format: pdf
date: 2022-10-14
toc: true
---

## Class 06

All functions in R have at least 3 things:

- A **name** (we pick this)
- Input **arguments** (there can be loads, comma separated)
- A **body** (the R code that does the)

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

# **Q1.**
**Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped.**

```{r}
#Define grade function
grade <- function(student) {
#Define student and Sort their grades
  student_sorted <- sort(student, decreasing=T, na.last=T)
  student_dropped <- head(student_sorted, n=-1)
  no_NA <- replace(student_dropped, is.na(student_dropped), 0)
  mean(no_NA)}
```

The overall grade of Student 1 is `r grade(student1)`

The overall grade of Student 2 is `r grade(student2)`

The overall grade of Student 3 is `r grade(student3)`

# **Q2.**
**Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook? [3pts]**

- Load the dataset
```{r}
gradebook <- read.csv("student_homework.csv", row.names=1)
gradebook_student_averages <- apply(gradebook,1, grade)
```
-Output the highest student average
The top student in the overall gradebook is Student `r which.max(gradebook_student_averages)`


# **Q3.**
**From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall? [2pts]**

-Convert NA to 0s for scores
```{r}
gradebook_no_NA <- replace(gradebook,is.na(gradebook),0)
```
-Create a function to calculate homework averages
```{r}
homework_difficulty <- function(homework) 
  {mean(homework)}
```
-Use the function to calculate homework averages
```{r}
gradebook_homework_averages <- apply(gradebook_no_NA[c(1:20),],2,homework_difficulty)
gradebook_homework_averages
```
-Output which homework had the lowest average
The toughest homework for students was Homework `r which.min(gradebook_homework_averages)` 

# **Q4.**

```{r}
mask <- gradebook
mask[ is.na(mask) ] <- 0

hw_scores <- apply(gradebook, 2, sum, na.rm=T)
hw_scores
```
The homework that students scored lowest on was `r which.min(hw_scores)`

**Q5. Make sure you save your Quarto document and can click the "Render" (or Rmark- down"Knit") button to generate a PDF foramt report without errors. Finally, submit your PDF to gradescope. [1pt]**
