---
title: "Class 10: Halloween Candy Mini-Project"
author: "Eric Jordahl"
format: html
date: 2022-10-28
toc: True
---

## Background

In this mini project we will examine 538 halloween candy data. 

```{r}
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"

candy <-  read.csv(url, row.names=1)
head(candy)
```

# **Q1.** 
**How many different candy types are in this dataset?**

```{r}
nrow(candy)
```
There are `r nrow(candy)` candies in this dataset

# **Q2.**
**How many fruity candy types are in the dataset?**

```{r}
sum(candy$fruity)
```

There are `r sum(candy$fruity)` fruity candies in this dataset

## What is your favorite Candy?

# **Q3**
**What is your favorite candy in the dataset and what is it’s winpercent value?**

```{r}
candy["Kit Kat", ]$winpercent
```

My favorite candy is Kit Kat and the win percentage is `r candy["Kit Kat", ]$winpercent`.

# **Q4**
**What is the winpercent value for “Kit Kat”?**

The win percent value for Kit Kat `r candy["Kit Kat", ]$winpercent`.

# **Q5**
**What is the winpercent value for Tootsie Roll Snack Bars?**

The win percent value for Tootsie Roll Snack Bars is `r candy["Tootsie Roll Snack Bars", ]$winpercent`.

```{r}
library("skimr")
skim(candy)
```

# **Q6.**
**Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?**

The win percentage average is much higher than most of the other columns as it is in percentage not fraction of the data that is chosen.

# **Q7** 
**What do you think a zero and one represent for the candy$chocolate column?**
The zero and one would represent either yes or no for if the candy is chocolate or not, 

# **Q8**
**Plot a histogram of winpercent values**

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)
ggplot(candy, aes(winpercent)) + geom_histogram(bins=10, fill = "orange", col="black")
```

# **Q9**
**Is the distribution of winpercent values symmetrical?**

No, the distribution is not symmetrical

# **Q10**
**Is the center of the distribution above or below 50%?**
The center of the distribution is below 50%, as the center/median is `r median(candy$winpercent)`

#**Q11**
**On average is the cholocate candy higher or lower ranked than fruit candy?**

```{r}
choc_mean_winper <- mean(candy$winpercent[as.logical(candy$chocolate)])
fruit_mean_winper <- mean(candy$winpercent[as.logical(candy$fruity)])
choc_mean_winper > fruit_mean_winper
```

# **Q12**
**Is this difference statistically significant?**

```{r}
choc_stats <- candy$winpercent[as.logical(candy$chocolate)]
fruit_stats <- candy$winpercent[as.logical(candy$fruity)]
t.test(choc_stats, fruit_stats)
```

Yes this is statistically significant becasue the p-value is less than .05

## Overall Candy Rankings

# **Q13**
**What are the five least liked candy types in this set?**

```{r}
library(dplyr)

candy %>% arrange(winpercent) %>% head(5)
```

The least liked candies are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters

# **Q14**
**What are the top 5 all time favorite candy types out of this set?**

```{r}
library (dplyr)

candy %>% arrange(winpercent) %>% tail(5)
```

The top 5 candies are Snickers, Kit Kat, Twix, Reese's Miniatures, and Reese's Peanut Butter Cup

# **Q15.**
**Make a first barplot of candy ranking based on winpercent values**

```{r}
ggplot(candy) + aes(winpercent, rownames(candy)) + geom_col() + theme(text=element_text(size=5))
```

# **Q16**
**This is quite ugly, use the reorder() function to get the bars sorted by winpercent?**

```{r}
ggplot(candy) + aes(winpercent, reorder(rownames(candy),winpercent)) + geom_col() + theme(text=element_text(size=5))
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + aes(winpercent, reorder(rownames(candy),winpercent), ) + geom_col(fill=my_cols) + theme(text=element_text(size=5))
```

# **Q17.**
**What is the worst ranked chocolate candy?**

The worst ranked chocolate candy is "Sixlets"

# **Q18**
**What is the best ranked fruity candy?**

The Best ranked fruity candy is "Starburst"

## Taking a look at Pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
 aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=2.5, max.overlaps = 15)
```

#**Q19** 
**Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?**

The candy that is ranked highest with the lowest price is the Reese's Miniatures as it has a high win percentage with an overall low price

# **Q20**
**What are the top 5 most expensive candy types in the dataset and of these which is the least popular?**

```{r}
library (dplyr)

most_exp_candy <-candy %>% arrange(pricepercent) %>% tail(5)
most_exp_candy_order <- most_exp_candy %>% arrange(winpercent) %>% head(1)
most_exp_candy_order
```

Nik L Nip has the lowest ranking (by winpercent) of the 5 most expensive candies.

# **Q21**
**Make a lollipop chart of the data by pricepercent vs candy**

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point() + theme(text=element_text(size=5))
```

## Exploring the Correlation Structure
```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

# **Q22**
**Examining this plot what two variables are anti-correlated (i.e. have minus values)?**

Chocolate and Fruity are the most negatively correlated of these categories

# **Q23**
**Similarly, what two variables are most positively correlated?**

Chocolate and winpercentage or bar are very positively correlated 

## Principle Component Analysis

```{r}
pca <- prcomp(candy, scale=T)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
      subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

# **Q24**
**What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?**

The original variables that are picked up are fruity, hard, and pluribus
Yes I would say this generally makes sense as fruity candies tend to have many in a package and they tend to be hard-er candies than many other candies. They are shown as also positively correlated in our correlation plot.